Type Refinements for the Working Class

### 2015-11-22 paper.tex
Restore equality of parametric refinements, functional refinement

\todo[inline]{%
  Restore definitions of the refinement framework after they have been fixed!!
}

%%
%% \subsection{Parametric Equality}
%%
%% The primary judgment concerning refinements is the parametric equality,
%% $\RMemEq{\Phi}{M}{N}{\phi}$, which presupposes $\RefinesCtx{\Phi}{\Upsilon}$,
%% $\Refines{\Phi}{\phi}{\tau}$, $\IsAbt{\cdot}{\Upsilon}{\cdot}{M}{\tau}$ and
%% $\IsAbt{\cdot}{\Upsilon}{\cdot}{N}{\tau}$. The meaning of this judgment is that
%% $M$ and $N$ are identified by $\phi$ at the identity renaming
%% $\Of{1}{\IMode{\Upsilon}\inj\OMode{\Upsilon}}$:
%% \[
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }{
%%     \IMode{M}\sim^1_\phi\IMode{N}
%%   }
%% \]
%%
%% \subsection{Functional Refinement}
%%
%% \newcommand\PRefinesCtxNil{\leftidx{^\parallel}{\!\!\sqsubset}{^\star_{\mathtt{nil}}}}
%% \newcommand\PRefinesCtxSnoc{\leftidx{^\parallel}{\!\!\sqsubset}{^\star_{\mathtt{snoc}}}}
%%
%% Next, we define functional refinement
%% $\HEqRefines{\Phi}{\Psi}{\phi}{\psi}{\tau}$ in terms of parametric refinement,
%% simultaneously with parametric context refinement
%% $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$ and functional equality
%% $\HRMemEq{\Phi}{\Psi}{M}{N}{\phi}$. We will write
%% $\HRefines{\Phi}{\Psi}{\phi}{\tau}$ as a shorthand for
%% $\HEqRefines{\Phi}{\Psi}{\phi}{\phi}{\tau}$.
%%
%% \newcommand\DefPRefinesCtxNil[1]{
%%   \infer[\PRefinesCtxNil]{
%%     \PRefinesCtx{#1}{\cdot}{\cdot}
%%   }{
%%   }
%% }
%% \newcommand\DefPRefinesCtxSnoc[6]{
%%   \infer[\PRefinesCtxSnoc]{
%%     \PRefinesCtx{#1}{#2,#3:#4}{#5,#3:#6}
%%   }{
%%     \PRefinesCtx{#1}{#2}{#5} &
%%     \HRefines{#1}{#2}{#4}{#6}
%%   }
%% }
%%
%% Parametric context refinement $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$ shall be
%% defined as follows, presupposing $\RefinesCtx{\Phi}{\Upsilon}$ and
%% $\IsVarCtx{\Gamma}$:
%% \begin{gather*}
%%   \DefPRefinesCtxNil{\Phi}\qquad
%%   \DefPRefinesCtxSnoc{\Phi}{\Psi}{x}{\phi}{\Gamma}{\tau}
%% \end{gather*}
%%
%% Functional refinement $\HEqRefines{\Phi}{\Psi}{\phi}{\psi}{\tau}$ presupposes
%% $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$ and $\IsSort{\tau}$, and is explained by
%% induction on the evidence for its first presupposition.
%%
%% \paragraph{Case} $\DefPRefinesCtxNil{\Phi}$.
%% \[
%%   \infer{
%%     \HEqRefines{\Phi}{\cdot}{\phi}{\psi}{\tau}
%%   }{
%%     \EqRefines{\Phi}{\phi}{\psi}{\tau}
%%   }
%% \]
%%
%% \paragraph{Case} $\DefPRefinesCtxSnoc{\Phi}{\Psi}{x}{\chi}{\Gamma}{\sigma}$,
%% where $\RefinesCtx{\Phi}{\Upsilon}$.
%%
%% \begin{gather*}
%%   \infer{
%%     \HEqRefines{\Phi}{\Psi,x:\chi}{\phi}{\psi}{\tau}
%%   }{
%%     \HPGenJ{\Upsilon}{\Gamma}{\MV{m},\MV{n}:\sigma}{
%%       \HypJ{
%%         \HEqRefines{\Phi}{\Psi}{
%%           \Subst{\MV{m}}{x}{\phi}
%%         }{
%%           \Subst{\MV{n}}{x}{\psi}
%%         }{\tau}
%%       }{
%%         \HRMemEq{\Phi}{\Psi}{\MV{n}}{\MV{n}}{\chi}
%%       }
%%     }
%%   }
%% \end{gather*}
%%
%% \subsection{Functional Equality}
%%
%% Functional equality $\HRMemEq{\Phi}{\Psi}{M}{N}{\phi}$ presupposes
%% $\RefinesCtx{\Phi}{\Upsilon}$, $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$,
%% $\HRefines{\Phi}{\Psi}{\phi}{\tau}$, $\IsAbt{\cdot}{\Upsilon}{\Gamma}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\Gamma}{N}{\tau}$; the judgment is defined by
%% induction on the evidence for the second presupposition
%% $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$:
%%
%% \paragraph{Case} $\DefPRefinesCtxNil{\Phi}$.
%% \[
%%   \infer{
%%     \HRMemEq{\Phi}{\cdot}{M}{N}{\phi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }
%% \]
%%
%% \paragraph{Case} $\DefPRefinesCtxSnoc{\Phi}{\Psi}{x}{\psi}{\Gamma}{\sigma}$,
%% where $\RefinesCtx{\Phi}{\Upsilon}$.
%% \[
%%   \infer{
%%     \HRMemEq{\Phi}{\Psi,x:\psi}{M}{N}{\phi}
%%   }{
%%     \HPGenJ{\Upsilon}{\Gamma}{\MV{m},\MV{n}:\sigma}{
%%       \HypJ{
%%         \HRMemEq{\Phi}{\Psi}{
%%           \Subst{\MV{m}}{x}{\phi}
%%         }{
%%           \Subst{\MV{n}}{x}{\psi}
%%         }{\phi}
%%       }{
%%         \HRMemEq{\Phi}{\Psi}{\MV{m}}{\MV{n}}{\psi}
%%       }
%%     }
%%   }
%% \]
%%
%% \section{Constructions on Refinements}
%%
%% Refinements can be arranged in a partial order via the
%% $\SubRefines{\Phi}{\phi}{\psi}{\tau}$ judgment, and admit a lattice structure
%% at any sort $\tau$.
%%
%% We have the bottom refinement $\GRefines{\TyBot{\tau}}{\tau}$, which simply
%% contains no elements. Next, the top refinement $\GRefines{\TyTop{\tau}}{\tau}$
%% identifies all terms $M,N$ such that $\IsAbt{\cdot}{\Upsilon}{\cdot}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\cdot}{N}{\tau}$. The join
%% $\Refines{\Phi}{\phi\TyJoin{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined as follows:
%% \begin{gather*}
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }\qquad
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \end{gather*}
%%
%% Finally the meet $\Refines{\Phi}{\phi\TyMeet{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined
%% dually:
%% \[
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyMeet{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi} &
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \]
%%
%% It is trivial to verify that these constructions on refinements give rise to a
%% lattice.
%%
%% \section{Multi-Sorted Computational Type Theory}
%%
%% Martin-L\"of's type theory is a theory of \emph{value types}: to define a type,
%% you specify how to form its equal canonical inhabitants, or values. Then, the
%% inhabitants of the types are explained uniformly by extending this relation
%% over an operational semantics such that two terms are equal members of a
%% type just when they compute to equal values of that type.
%%
%% More generally, one may wish to consider types which include computations in
%% addition to values; an example of such a type would be $\bar{\psi}$ when $\psi$
%% is a type, which would include all the members of $\psi$ as well as a divergent
%% computation. Another example is the type $\TyTop{\tau}$ which identifies all
%% closed terms of sort $\tau$.
%%
%% In order to ensure that such types are sensible, we must formulate a notion of
%% computational equivalence, and then say that a refinement is a \emph{type} just
%% when it respects computational equivalence; then, we are able to define types
%% as PERs on \emph{programs} in addition to values.

>>>>

\subsection{Parametric Equality}

The primary judgment concerning refinements is the parametric equality,
$\RMemEq{\Upsilon}{M}{N}{\phi}$, which presupposes
$\Refines{\Upsilon}{\phi}{\tau}$, and $\Member{M,N}{\Exprs_\tau(\Upsilon)}$.
The meaning of this judgment is that $M$ and $N$ are identified by $\phi$ at $\Upsilon$:
\[
  \infer{
    \RMemEq{\Upsilon}{M}{N}{\phi}
  }{
    \RefEquate{\phi}{1}{M}{N}
  }
\]

\subsection{Functional Refinement}

\newcommand\PRefinesCtxNil{\leftidx{^\parallel}{\!\!\sqsubset}{^\star_{\mathtt{nil}}}}
\newcommand\PRefinesCtxSnoc{\leftidx{^\parallel}{\!\!\sqsubset}{^\star_{\mathtt{snoc}}}}

Next, we define functional refinement
$\HEqRefines{\Upsilon}{\Psi}{\phi}{\psi}{\tau}$ in terms of parametric refinement,
simultaneously with parametric context refinement
$\PRefinesCtx{\Upsilon}{\Psi}{\Gamma}$ and functional equality
$\HRMemEq{\Upsilon}{\Psi}{M}{N}{\phi}$. We will write
$\HRefines{\Upsilon}{\Psi}{\phi}{\tau}$ as a shorthand for
$\HEqRefines{\Upsilon}{\Psi}{\phi}{\phi}{\tau}$.

\newcommand\DefPRefinesCtxNil[1]{
  \infer[\PRefinesCtxNil]{
    \PRefinesCtx{#1}{\cdot}{\cdot}
  }{
  }
}
\newcommand\DefPRefinesCtxSnoc[6]{
  \infer[\PRefinesCtxSnoc]{
    \PRefinesCtx{#1}{#2,#3:#4}{#5,#3:#6}
  }{
    \PRefinesCtx{#1}{#2}{#5} &
    \HRefines{#1}{#2}{#4}{#6}
  }
}

Parametric context refinement $\PRefinesCtx{\Upsilon}{\Psi}{\Gamma}$ shall be
defined as follows, presupposing $\IsSymCtx{\Upsilon}$ and
$\IsVarCtx{\Gamma}$:
\begin{gather*}
  \DefPRefinesCtxNil{\Upsilon}\qquad
  \DefPRefinesCtxSnoc{\Upsilon}{\Psi}{x}{\phi}{\Gamma}{\tau}
\end{gather*}

Functional refinement $\HEqRefines{\Upsilon}{\Psi}{\phi}{\psi}{\tau}$ presupposes
$\PRefinesCtx{\Upsilon}{\Psi}{\Gamma}$ and $\IsSort{\tau}$, and is explained by
induction on the evidence for its first presupposition.

\paragraph{Case} $\DefPRefinesCtxNil{\Upsilon}$.
\[
  \infer{
    \HEqRefines{\Upsilon}{\cdot}{\phi}{\psi}{\tau}
  }{
    \EqRefines{\Upsilon}{\phi}{\psi}{\tau}
  }
\]

\paragraph{Case} $\DefPRefinesCtxSnoc{\Upsilon}{\Psi}{x}{\chi}{\Gamma}{\sigma}$.

\begin{gather*}
  \infer{
    \HEqRefines{\Upsilon}{\Psi,x:\chi}{\phi}{\psi}{\tau}
  }{
    \HPGenJ{\Upsilon}{\Gamma}{\MV{m},\MV{n}:\sigma}{
      \HypJ{
        \HEqRefines{\Upsilon}{\Psi}{
          \Subst{\MV{m}}{x}{\phi}
        }{
          \Subst{\MV{n}}{x}{\psi}
        }{\tau}
      }{
        \HRMemEq{\Upsilon}{\Psi}{\MV{n}}{\MV{n}}{\chi}
      }
    }
  }
\end{gather*}

\todo[inline]{%
  Is the above a correct rule? I think that it might be good to be explicit
  here rather than using the hypothetico-general judgment. In fact, if we are
  explicit about everything, we can get away with not even explaining the
  higher-order judgments.
}

%%
%% \subsection{Functional Equality}
%%
%% Functional equality $\HRMemEq{\Phi}{\Psi}{M}{N}{\phi}$ presupposes
%% $\RefinesCtx{\Phi}{\Upsilon}$, $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$,
%% $\HRefines{\Phi}{\Psi}{\phi}{\tau}$, $\IsAbt{\cdot}{\Upsilon}{\Gamma}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\Gamma}{N}{\tau}$; the judgment is defined by
%% induction on the evidence for the second presupposition
%% $\PRefinesCtx{\Phi}{\Psi}{\Gamma}$:
%%
%% \paragraph{Case} $\DefPRefinesCtxNil{\Phi}$.
%% \[
%%   \infer{
%%     \HRMemEq{\Phi}{\cdot}{M}{N}{\phi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }
%% \]
%%
%% \paragraph{Case} $\DefPRefinesCtxSnoc{\Phi}{\Psi}{x}{\psi}{\Gamma}{\sigma}$,
%% where $\RefinesCtx{\Phi}{\Upsilon}$.
%% \[
%%   \infer{
%%     \HRMemEq{\Phi}{\Psi,x:\psi}{M}{N}{\phi}
%%   }{
%%     \HPGenJ{\Upsilon}{\Gamma}{\MV{m},\MV{n}:\sigma}{
%%       \HypJ{
%%         \HRMemEq{\Phi}{\Psi}{
%%           \Subst{\MV{m}}{x}{\phi}
%%         }{
%%           \Subst{\MV{n}}{x}{\psi}
%%         }{\phi}
%%       }{
%%         \HRMemEq{\Phi}{\Psi}{\MV{m}}{\MV{n}}{\psi}
%%       }
%%     }
%%   }
%% \]
%%
%% \section{Constructions on Refinements}
%%
%% Refinements can be arranged in a partial order via the
%% $\SubRefines{\Phi}{\phi}{\psi}{\tau}$ judgment, and admit a lattice structure
%% at any sort $\tau$.
%%
%% We have the bottom refinement $\GRefines{\TyBot{\tau}}{\tau}$, which simply
%% contains no elements. Next, the top refinement $\GRefines{\TyTop{\tau}}{\tau}$
%% identifies all terms $M,N$ such that $\IsAbt{\cdot}{\Upsilon}{\cdot}{M}{\tau}$
%% and $\IsAbt{\cdot}{\Upsilon}{\cdot}{N}{\tau}$. The join
%% $\Refines{\Phi}{\phi\TyJoin{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined as follows:
%% \begin{gather*}
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi}
%%   }\qquad
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyJoin{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \end{gather*}
%%
%% Finally the meet $\Refines{\Phi}{\phi\TyMeet{\tau}\psi}{\tau}$ of two refinements
%% $\Refines{\Phi}{\phi}{\tau}$ and $\Refines{\Phi}{\psi}{\tau}$ is defined
%% dually:
%% \[
%%   \infer{
%%     \RMemEq{\Phi}{M}{N}{\phi\TyMeet{\tau}\psi}
%%   }{
%%     \RMemEq{\Phi}{M}{N}{\phi} &
%%     \RMemEq{\Phi}{M}{N}{\psi}
%%   }
%% \]
%%
%% It is trivial to verify that these constructions on refinements give rise to a
%% lattice.
%%
%% \section{Multi-Sorted Computational Type Theory}
%%
%% Martin-L\"of's type theory is a theory of \emph{value types}: to define a type,
%% you specify how to form its equal canonical inhabitants, or values. Then, the
%% inhabitants of the types are explained uniformly by extending this relation
%% over an operational semantics such that two terms are equal members of a
%% type just when they compute to equal values of that type.
%%
%% More generally, one may wish to consider types which include computations in
%% addition to values; an example of such a type would be $\bar{\psi}$ when $\psi$
%% is a type, which would include all the members of $\psi$ as well as a divergent
%% computation. Another example is the type $\TyTop{\tau}$ which identifies all
%% closed terms of sort $\tau$.
%%
%% In order to ensure that such types are sensible, we must formulate a notion of
%% computational equivalence, and then say that a refinement is a \emph{type} just
%% when it respects computational equivalence; then, we are able to define types
%% as PERs on \emph{programs} in addition to values.
